---
title: "Models"
author: "Xue Qin"
date: "2024-10-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(caret)
library(pROC)
```

```{r}
data <- read_csv("../data/raw-data/aids_clinical_trials_combined.csv")
```
# Logistic Regression
```{r}
binary_cols <- c("hemo", "homo", "drugs", "karnof", "oprior", "z30", "zprior", 
                "preanti", "race", "gender", "str2", "strat", "symptom", 
                "treat", "offtrt")

# Survival variables (don't scale these)
survival_cols <- c("time", "cid")

# Get all column names
all_cols <- names(data)

# Identify continuous columns to scale
# (those not in binary_cols and not in survival_cols)
continuous_cols <- setdiff(all_cols, c(binary_cols, survival_cols))

# Create new dataframe
data_scaled <- data

# Scale only continuous variables
data_scaled[continuous_cols] <- scale(data[continuous_cols])

# Binary and survival variables remain unchanged
data_scaled[binary_cols] <- data[binary_cols]
data_scaled[survival_cols] <- data[survival_cols]
data_scaled_950<- data_scaled%>%
  filter(time<=950)

write.csv(data_scaled_950,"../data/derived-data/data_scaled_950.csv",row.names = FALSE)

library(caret)

set.seed(123)

trainIndex <- createDataPartition(data_scaled_950$cid, p = 0.7, list = FALSE)

train_data <- data_scaled_950[trainIndex, ]
test_data <- data_scaled_950[-trainIndex, ]
train_data$cid <- as.factor(train_data$cid)
test_data$cid <- as.factor(test_data$cid)
```

```{r}
logistic_model <- glm(cid ~ age + trt + wtkg + hemo + homo + drugs + karnof+oprior+z30+zprior+preanti+race+gender+str2+strat+symptom+treat+offtrt+cd40+cd420+cd80+cd820, 
                   data = train_data, family = binomial)
model_summary<- summary(logistic_model)
```

```{r}
test_predictions <- predict(logistic_model, newdata = test_data, type = "response")

test_pred_class <- ifelse(test_predictions > 0.5, 1, 0)

confusionMatrix(factor(test_pred_class), factor(test_data$cid))

```

```{r}
# Roc Curve

roc_curve <- roc(test_data$cid, test_predictions)
plot(roc_curve, col = "blue", main = "ROC Curve for Logistic Regression")

auc(roc_curve)

```

```{r}
p_values <- model_summary$coefficients[, 4] 

significant_vars <- names(p_values[p_values < 0.05])

print(significant_vars)
```

## fit with selected features with P<0.05
```{r}
logistic_model2 <- glm(cid ~ karnof+preanti+symptom+treat+cd40+cd420+cd820, 
                   data = train_data, family = binomial)

# Summary of the model
model_summary2<- summary(logistic_model2)
```

```{r}

test_predictions2 <- predict(logistic_model2, newdata = test_data, type = "response")

test_pred_class2 <- ifelse(test_predictions2 > 0.5, 1, 0)

confusionMatrix(factor(test_pred_class2), factor(test_data$cid))
```

# Lasso
```{r}
library(glmnet)


X <- model.matrix(cid ~ age + trt + wtkg + hemo + homo + drugs + karnof + oprior + z30 + zprior + preanti + race + gender + str2 + strat + symptom + treat + offtrt + cd40 + cd420 + cd80 + cd820, data = data_scaled_950)[, -1]
y <- data_scaled_950$cid

lasso_model <- glmnet(X, y, family = "binomial", alpha = 1)


cv_lasso <- cv.glmnet(X, y, family = "binomial", alpha = 1)


best_lambda <- cv_lasso$lambda.min

coef(cv_lasso, s = "lambda.min")

```

## fit model with selected feature
```{r}

X_train <- model.matrix(cid ~ offtrt + cd420 + race + hemo + symptom + drugs + gender + cd820 + age + strat, data = train_data)[, -1]
X_test <- model.matrix(cid ~ offtrt + cd420 + race + hemo + symptom + drugs + gender + cd820 + age + strat, data = test_data)[, -1]


y_train <- train_data$cid
y_test <- test_data$cid
cv_lasso <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1)

best_lambda <- cv_lasso$lambda.min

lasso_model <- glmnet(X_train, y_train, family = "binomial", alpha = 1, lambda = best_lambda)

test_predictions_lasso <- predict(lasso_model, newx = X_test, type = "response")

roc_curve_lasso <- roc(test_data$cid, test_predictions_lasso)
plot(roc_curve_lasso, main = "ROC Curve for Lasso Selection Model", col="violetred2")
auc(roc_curve_lasso)

test_pred_class_lasso <- ifelse(test_predictions_lasso > 0.5, 1, 0)
# Confusion Matrix
confusionMatrix(factor(test_pred_class_lasso), factor(y_test))

print(coef(lasso_model))

```


### Backward Selection

```{r}
# Perform backward selection
backward_model <- step(logistic_model, direction = "backward", trace = FALSE)

backward_model_summary <- summary(backward_model)

test_predictions_backward <- predict(backward_model, newdata = test_data, type = "response")
test_pred_class_backward <- ifelse(test_predictions_backward > 0.5, 1, 0)

conf_matrix <- confusionMatrix(factor(test_pred_class_backward), factor(test_data$cid))
print("Confusion Matrix and Performance Metrics for Backward Selection Model:")
print(conf_matrix)

# Get the selected features
selected_features <- names(coef(backward_model))
print("Features Selected by Backward Selection:")
print(selected_features)

# Plot ROC curve for backward selection model
roc_curve_backward <- roc(test_data$cid, test_predictions_backward)
plot(roc_curve_backward, main = "ROC Curve for Backward Selection Model", col="orange")
auc(roc_curve_backward)

# Print coefficients of the final model
print("Coefficients of the Final Model:")
print(coef(backward_model))
```

```{r}
plot(roc_curve_backward, col = "orange", main = "Comparison of ROC Curves",
     xlab = "False Positive Rate (1 - Specificity)", 
     ylab = "True Positive Rate (Sensitivity)")

lines(roc_curve_lasso,col="violetred2")
# Add the second ROC curve
lines(roc_curve, col = "lightblue")

# Add diagonal reference line
abline(a = 0, b = 1, lty = 2, col = "gray")

# Add legend
legend("bottomright", 
       legend = c(paste("Backward Selection (AUC =", round(auc(roc_curve_backward), 4), ")"),
                 paste("Lasso Model (AUC =", round(auc(roc_curve_lasso), 4), ")"),
                 paste("Full Model (AUC =", round(auc(roc_curve), 4), ")")),
       col = c("orange", "violetred2","lightblue"), 
       lwd = 2)
```

```{r}
selected_features <- c("offtrt", "cd420", "race", "hemo", "symptom", "drugs", "gender", "cd820", "age", "strat")
formula <- as.formula(paste("cid ~", paste(selected_features, collapse = " + ")))

logistic_model <- glm(formula, data = train_data, family = binomial)
summary(logistic_model)
```

